Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-12 latex/Abstract/Abstract.tex
update

\noindent
\textbf{Motivation}
\label{Abstract}
\\
Reinforcement Learning (\textbf{RL}), exposed with brilliant clarity in \cite{Sutton1998}, has until recently known less success than we might have hoped for. Its framework is very appealing and intuitive. In particular, the mathematical beauty of Value iteration and Q iteration (\cite{WatkinsThesis}) for discrete state and action spaces, blindly iterating from \textit{any} initial value, is quite profound. Sadly, it had until recently proven hard to achieve practical success with these methods.
\\
Inspired however by the seminal success of Deep Mind's team in using Deep Reinforcement Learning (\textbf{DRL}) to play Atari games (\cite{Mnih2013})  and to master the game of Go (\cite{AlphaGo}), researchers have in recent years made a lot of progress towards designing algorithms capable of learning and solving, \textit{without human knowledge}, the Rubik's Cube (\textbf{RC}) - as well as similar single player puzzles - by using Deep Q-Learning (\textbf{DQL})  (\cite{DBLP:journals/corr/abs-1805-07470}) or search and (\textbf{DRL}) value iteration (\cite{https://doi.org/10.48550/arxiv.1805.07470}).
\\
In this project, I will attempt to implement a variety of solvers combining $A^{*}$ search and heuristics, some of which will be handcrafted, others which I will train on randomly generated sequences of puzzles using \textbf{DL}, \textbf{DRL} or \textbf{DQL}, to solve the 15-puzzle (and variations of different dimensions), as well as the Rubik's cube.
\\
\\
\textbf{Organisation of this thesis}
\\
In the first section, I will quickly describe what I hope to get out of this project, both in terms of personal learning, as well as in terms of tangible results (solving some puzzles!). In section 2, I will do a quick recap of the different methods that I will use in the project. Section 3 will be dedicated to discussing the mathematics of the sliding puzzle \textbf{SP} and the \textbf{RC}. I might throw a few random (but hopefully interesting) observations in there and give some references for the keen reader. I will then give an overview in section 4 of the code base I have developed - and put in the open on my github page (\cite{FB}) - to complete this project. Finally, in sections 6 and 7, I will present all my various results on respectively the sliding puzzle and the Rubik's cube before offering some conclusion in section 8.
\\
A couple of appendix sections complement the main text. The first one details how to use the code base, from constructing puzzles, to learning and solving them using all the different methods I have implemented for this project. The second describes a limitation I bumped into while using the kociemba library (a well-known open source 3x3x3 \textbf{RC} solver), and how I circumvented it.

>>>>

\label{Abstract}
\noindent \textbf{Motivation} -- Reinforcement Learning (\textbf{RL}), exposed with brilliant clarity in \cite{Sutton1998}, has until recently known less success than we might have hoped for. Its framework is very appealing and intuitive. In particular, the mathematical beauty of Value iteration and Q iteration (\cite{WatkinsThesis}) for discrete state and action spaces, blindly iterating from \textit{any} initial value, is quite profound. Sadly, it had until recently proven hard to achieve practical success with these methods. Inspired however by the seminal success of Deep Mind's team in using Deep Reinforcement Learning (\textbf{DRL}) to play Atari games (\cite{Mnih2013})  and to master the game of Go (\cite{AlphaGo}), researchers have in recent years made a lot of progress towards designing algorithms capable of learning and solving, \textit{without human knowledge}, the Rubik's Cube (\textbf{RC}) - as well as similar single player puzzles - by using Deep Q-Learning (\textbf{DQL})  (\cite{DBLP:journals/corr/abs-1805-07470}) or search and (\textbf{DRL}) value iteration (\cite{https://doi.org/10.48550/arxiv.1805.07470}).
\\
\\
In this project, I will attempt to implement a variety of solvers combining $A^{*}$ search and heuristics, some of which will be handcrafted, others which I will train on randomly generated sequences of puzzles using \textbf{DL}, \textbf{DRL} or \textbf{DQL}, to solve the 15-puzzle (and variations of different dimensions), as well as the Rubik's cube.
\\
\\
\textbf{Organisation of this thesis} -- In the first section, I describe what I hope to get out of this project, both in terms of personal learning, as well as in terms of tangible results (solving some puzzles!). In section 2, I will do a recap of the different methods used in the project. Section 3 is dedicated to discussing the mathematics of the sliding puzzle \textbf{SP} and the \textbf{RC}. I might make interesting observations in there and give some references for the keen reader. I give then an overview in section 4 of the code base I have developed (see my github page \cite{FB}) - for this project. Finally, sections 6 and 7 present my results on respectively the \textbf{SP} and the \textbf{RC} and section 8 concludes the thesis.
\\
\\
A couple of appendix sections complement the main text. The first one details how to use the code base, from constructing puzzles, to learning and solving them using the various solvers I have developed. The second one describes a limitation I bumped into while using the kociemba library (a well-known open source 3x3x3 \textbf{RC} solver), and how I circumvented it.
\newpage
\thispagestyle{empty}
\noindent \textbf{Acknowledgements} -- 
\input{Acknowledgements/Acknowledgements}

### 2022-09-12 latex/Chapters/Rubiks.tex
update

\Subsection{Deep Reinforcement Learner \& Deep Q Learner}
\label{sec:S33DRL}
To get my \textbf{DRL} and \textbf{DQL} heuristics, I have run the DeepReinforcementLearner and DeepQLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency=1000}) or when the MSE falls below 10 basis points of the maximum target cost-to-go (\textit{update\_target\_network\_threshold=1e-3})
\item Run for a maximum of 11,000 epochs (\textit{nb\_epochs=11000}) or when the maximum target so far has not been increasing by more than 1\% in the last 5,500 epochs (\textit{max\_target\_uptick=1e-2} and \textit{max\_target\_not\_increasing\_epochs\_pct=0.5})
\item At every update of the target network (\textit{training\_data\_every\_epoch=False}), generate 10,00 sequences of puzzles (\textit{nb\_sequences=10,000}), each comprised of a 15-scramble puzzle back to goal state, with all intermediary puzzles along the path (\textit{nb\_shuffles=15}).
\item Use a fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers of size 600, 300, 100 (\textit{layers\_description=(600,300,100)}) and use torch.optim.RMSProp as optimiser (\textit{optimiser=rms\_prop}) with an exponential scheduler (\textit{scheduler=exponential\_scheduler}) with gamma 0.999 (\textit{gamma\_scheduler=0.9999}) and \textit{learning\_rate=0.001}. The networks architecture is shown in figure \ref{fig:222RCnets}.
\end{itemize}
Graphs \ref{fig:222RCDRL} and \ref{fig:222RCDQL} show both learners' following metrics tracked over the epochs:
\begin{itemize}
\item learning rate, which decreases due to the scheduler, but gets reset at each network update.
\item max target. One interesting dynamic during learning is that due to the way targets are constructed using value-iteration update, and starting from an all 0 network, the max target only increases by about 1 at best at every network update. That is, the first network only learns how to differentiate goal from non-goal, then the second one learns how to differentiate goals, cost 1 and cost 2+ in some sense. This is why I have introduced the first set of parameters discussed just above. The first one makes sure we do not waste time initially training the left-hand-side network when training is easy (as consist of just goal versus not goal). The second ones make sure that we stop training altogether when the max target does not grow anymore. This is because we do not know a-priori what the God number of a given puzzle/dimension is, but let's assume for the sake of the argument that it is 15. Then that means that after 15 updates of the target network we might not really get so much benefit by keeping training and value-iterating.
\item the loss (MSE for \textbf{DRL}, MSE + CrossEntropyLoss for \textbf{DQL})
\item the loss divided by the max target. The exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 0.01 means little if the we do not compare it to the possible range that the cost-to-go can take.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the DeepReinforcementLearner. This is an interesting quantity, as it tells us later on when testing out of sample and comparing with other solvers, how well the solvers are able to generalise. In this case (3x3) we can see that the DeepReinforcementLearner quickly has seen a very large percentage of all puzzles. That being said, it is good to keep in mind that \textit{seen} does not mean really much here, since the learning is unsupervised. This is quite in contrast to the DeepLearner, for which puzzles seen come along with the cost-to-go computed by a \textit{teacher}, usually an optimal or at least efficient solver.
\end{itemize}

@@

We summarize as usual the 2x2x2 \textbf{RC} performance tests by our usual metrics \ref{fig:222RCPerformance}. Overall, the \textbf{DL} and \textbf{DRL} managed to solve all cubes (though \textbf{DL} took over an houd and \textbf{DRL} up to 6 hours for the slowest cube they encountered. Both methods outperformed in terms of optimality score Kociemba and got on par with A$^{*}[Kociemba]$. Their max cost over all cubes encountered was 12 for up to 20 shuffles, and 13 for $\infty$ shuffling. Considering God's number for the 2x2x2 \textbf{RC} of 11, this is rather impressive!
\\
\\
An very interesting remark is that if we check the learning graph \ref{fig:222RCDRL} of the \textbf{DRL}, we can see it only ever assigned a cost-to-go (value function) of 7.6 at most. Despite that fact, it was still able to solve cubes requiring 13 moves. How is that possible? Notice that the loss of 0.2 towards which it converged at the end of its training is misleading, since it is not a loss versus the actual true cost-to-go, but rather versus that of the target network. My theory is that while the value the \textbf{DRL} assigned to hard-to-solve cubes is off (by at least 4 or 5), it is likely the case that the network gets the relative value of neighbouring cubes about right, and this is sufficient to guide A$^{*}$ extremely well.
\\
\\
Finally, we can see that \textbf{DQL}, despite showing seemingly/visually cleaner and more monotonic loss convergence than \textbf{DRL} (see \ref{fig:222RCDQL} versus \ref{fig:222RCDRL}), struggled a lot more to solve difficult cubes (above 10 shuffles). Clearly this shows that my earlier findings on the 3x3 \textbf{SP} are likely not universal, and very much depend on the parameters set during training, as well as on the problem at hand.

>>>>

\Subsection{Deep Reinforcement Learner \& Deep Q Learner}
\label{sec:S33DRL}
To get my \textbf{DRL} and \textbf{DQL} heuristics, I have run the DeepReinforcementLearner and DeepQLearner with the following main parameters:
\begin{itemize}
\item Update the target network every 1,000 epochs (\textit{update\_target\_network\_frequency=1000}) or when the MSE falls below 10 basis points of the maximum target cost-to-go (\textit{update\_target\_network\_threshold=1e-3})
\item Run for a maximum of 11,000 epochs (\textit{nb\_epochs=11000}) or when the maximum target so far has not been increasing by more than 1\% in the last 5,500 epochs (\textit{max\_target\_uptick=1e-2} and \textit{max\_target\_not\_increasing\_epochs\_pct=0.5})
\item At every update of the target network (\textit{training\_data\_every\_epoch=False}), generate 10,00 sequences of puzzles (\textit{nb\_sequences=10,000}), each comprised of a 15-scramble puzzle back to goal state, with all intermediary puzzles along the path (\textit{nb\_shuffles=15}).
\item Use a fully connected network (\textit{network\_type=fully\_connected\_net}) with three hidden layers of size 600, 300, 100 (\textit{layers\_description=(600,300,100)}) and use torch.optim.RMSProp as optimiser (\textit{optimiser=rms\_prop}) with an exponential scheduler (\textit{scheduler=exponential\_scheduler}) with gamma 0.999 (\textit{gamma\_scheduler=0.9999}) and \textit{learning\_rate=0.001}. The networks architecture is shown in figure \ref{fig:222RCnets}.
\end{itemize}
Graphs \ref{fig:222RCDRL} and \ref{fig:222RCDQL} show both learners' following metrics tracked over the epochs:
\begin{itemize}
\item learning rate, which decreases due to the scheduler, but gets reset at each network update.
\item max target. One interesting dynamic during learning is that due to the way targets are constructed using value-iteration update, and starting from an all 0 network, the max target only increases by about 1 at best at every network update. That is, the first network only learns how to differentiate goal from non-goal, then the second one learns how to differentiate goals, cost 1 and cost 2+ \dots. This is why I have introduced the \textit{update\_target\_network\_threshold} discussed above. It makes sure we do not waste time initially training the left-hand-side network when training is easy. \textit{max\_target\_uptick} and \textit{max\_target\_not\_increasing\_epochs\_pct} on the other hand make sure that we stop training altogether when the max target does not grow anymore. This is because we do not know a-priori what the God number of a given puzzle/dimension is. Let us assume for the sake of the argument that it is 15: that means that beyond 15 updates of the target network we might not get much benefit to additional training and value-iterating.
\item the loss (MSE for \textbf{DRL}, MSE + CrossEntropyLoss for \textbf{DQL})
\item the loss divided by the max target. The exit criterion \textit{update\_target\_network\_threshold} is based on that quantity, since a loss of e.g. 0.01 means little if the we do not compare it to the possible range that the cost-to-go can take.
\item The percentage of all the possible puzzles of that dimension \textit{seen} by the learner. This is an interesting quantity, as it tells us how well the solvers are able to generalise. From both graphs (\ref{fig:222RCDRL} \& \ref{fig:222RCDQL}) we can see that the learners saw less than 1\% of all puzzles. It is good to keep in mind however that \textit{seen} does not mean really much here, since the learning is \textit{unsupervised}. This is quite in contrast to the DeepLearner, whose traiing data consist of puzzles with their cost-to-go computed by a \textit{teacher} (optimal or other efficient solver).
\end{itemize}

@@

We summarize as usual the 2x2x2 \textbf{RC} performance tests by our usual metrics \ref{fig:222RCPerformance}. Overall, the \textbf{DL} and \textbf{DRL} managed to solve all cubes (though \textbf{DL} took over an houd and \textbf{DRL} up to 6 hours for the slowest cube they encountered. Both methods outperformed in terms of optimality score Kociemba and got on par with A$^{*}[Kociemba]$. Their max cost over all cubes encountered was 12 for up to 20 shuffles, and 13 for $\infty$ shuffling. Considering God's number for the 2x2x2 \textbf{RC} of 11, this is rather impressive!
\\
\\
An very interesting remark is that if we check the learning graph \ref{fig:222RCDRL} of the \textbf{DRL}, we can see it only ever assigned a cost-to-go (value function) of 7.6 at most. Despite that fact, it was still able to solve cubes requiring 13 moves. How is that possible? Notice that the loss of 0.2 towards which it converged at the end of its training is misleading, since it is not a loss versus the actual true cost-to-go, but rather versus that of the target network. My theory is that while the value the \textbf{DRL} assigned to hard-to-solve cubes is off (by at least 4 or 5), it is likely the case that the network gets the relative value of neighbouring cubes about right (given that the value-iteration teaches the solver that, it is not super surprising either), and this is sufficient to guide A$^{*}$ extremely well. If both the absolute value of the cost-to-go and the relative (among neighbouring cubes) was totally off, we would indeed not expect the heuristic to make A$^{*}$ any better than, say, \textbf{BFS} and we would therefore not expect it to be able to solve \textit{costly} cubes in 10-13 moves.
\\
\\
Finally, we can see that \textbf{DQL}, despite showing seemingly/visually cleaner and more monotonic loss convergence than \textbf{DRL} (see \ref{fig:222RCDQL} versus \ref{fig:222RCDRL}), struggled a lot more to solve difficult cubes (above 10 shuffles). Clearly this shows that my earlier findings on the 3x3 \textbf{SP} are likely not universal, and very much depend on the parameters set during training, as well as on the problem at hand.
